---
title: "Derek C. Briggs — CV (Short)"
author: "Derek C. Briggs"
date: last-modified
date-format: "MMMM YYYY"
format:
  html:
    theme: none
    css:
      - cv.css
      - cv-short.css
    toc: false
    anchor-sections: false
    link-external-newwindow: true
    include-in-header:
      text: |
        <meta name="description" content="Two-page curriculum vitae of Derek C. Briggs, Professor of Research and Evaluation Methodology, University of Colorado Boulder.">
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(yaml)
library(readr)

pubs_raw     <- yaml::read_yaml("data/publications.yaml")$publications
grants_raw   <- readr::read_csv("data/grants.csv", show_col_types = FALSE)

esc <- function(x) {
  if (is.null(x)) return("")
  x <- as.character(x)
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x
}

year_sort_key <- function(y) {
  if (is.numeric(y)) return(as.numeric(y))
  if (identical(y, "in press")) return(9999)
  if (identical(y, "in preparation")) return(9998)
  suppressWarnings(as.numeric(y))
}

vol_tail <- function(p) {
  out <- ""
  if (!is.null(p$volume)) {
    out <- esc(p$volume)
    if (!is.null(p$issue)) out <- paste0(out, "(", esc(p$issue), ")")
  }
  if (!is.null(p$pages)) {
    out <- if (nchar(out)) paste0(out, ", ", esc(p$pages)) else esc(p$pages)
  }
  if (nchar(out)) paste0(", ", out) else ""
}

loc_pub <- function(p) {
  if (is.null(p$publisher) && is.null(p$location)) return("")
  if (!is.null(p$location) && !is.null(p$publisher)) {
    sprintf("%s: %s", esc(p$location), esc(p$publisher))
  } else if (!is.null(p$publisher)) {
    esc(p$publisher)
  } else {
    esc(p$location)
  }
}

title_end <- function(p) {
  last <- substr(p$title, nchar(p$title), nchar(p$title))
  if (last %in% c("?", "!", ".")) "" else "."
}

link_html <- function(p) {
  if (!is.null(p$doi)) {
    sprintf(' <a href="https://doi.org/%s" class="pub-doi">doi.org/%s</a>',
            esc(p$doi), esc(p$doi))
  } else if (!is.null(p$url)) {
    sprintf(' <a href="%s" class="pub-doi">link</a>', esc(p$url))
  } else ""
}

format_pub <- function(p) {
  year_str <- as.character(p$year)
  t <- esc(p$title)
  te <- title_end(p)
  a <- esc(p$authors)

  body <- switch(p$category,
    "book" = {
      lp <- loc_pub(p)
      sprintf("%s (%s). <em>%s</em>%s%s",
              a, year_str, t, te,
              if (nchar(lp)) paste0(" ", lp, ".") else "")
    },
    "journal-article" = {
      venue <- if (!is.null(p$venue))
        sprintf(" <em>%s</em>%s.", esc(p$venue), vol_tail(p)) else ""
      sprintf("%s (%s). %s%s%s", a, year_str, t, te, venue)
    },
    "book-chapter" = {
      ed <- if (!is.null(p$editors)) paste0(" In ", esc(p$editors), ",") else " In"
      book <- if (!is.null(p$venue)) sprintf(" <em>%s</em>", esc(p$venue)) else ""
      pp <- if (!is.null(p$pages)) sprintf(" (pp. %s)", esc(p$pages)) else ""
      lp <- loc_pub(p); lp_str <- if (nchar(lp)) paste0(" ", lp, ".") else ""
      sprintf("%s (%s). %s%s%s%s%s.%s",
              a, year_str, t, te, ed, book, pp, lp_str)
    },
    sprintf("%s (%s). %s%s", a, year_str, t, te)
  )

  body <- paste0(body, link_html(p))
  sprintf('<div class="pub-entry">%s</div>', body)
}

render_featured <- function() {
  entries <- Filter(function(p) isTRUE(p$featured), pubs_raw)
  if (!length(entries)) return(invisible())
  keys <- vapply(entries, function(p) year_sort_key(p$year), numeric(1))
  entries <- entries[order(keys, decreasing = TRUE)]
  base::cat(paste(vapply(entries, format_pub, character(1)), collapse = "\n\n"))
  base::cat("\n")
}

# Grants summary numbers.
g_total <- sum(grants_raw$amount)
g_pi    <- sum(grants_raw$amount[grants_raw$role == "PI"])
g_copi  <- g_total - g_pi
g_n     <- nrow(grants_raw)
g_pi_n  <- sum(grants_raw$role == "PI")

fmt_m <- function(x) {
  if (x >= 1e6) sprintf("$%.2fM", x / 1e6) else sprintf("$%sK", formatC(round(x / 1e3), big.mark = ","))
}
```

::: {.dashboard}

::: {.dashboard-header}

[Full CV](cv.html){.full-cv-btn}

# Derek C. Briggs {.name}

::: {.titles}
Professor · Research and Evaluation Methodology · University of Colorado Boulder

Associate Dean of Faculty, School of Education · Director, CADRE
:::

::: {.contact}
[derek.briggs@colorado.edu](mailto:derek.briggs@colorado.edu) ·
[Website](https://derekcbriggs.github.io/) ·
[Google Scholar](https://scholar.google.com/citations?user=FsarJmwAAAAJ&hl=en) ·
[ORCID: 0000-0003-1628-4661](https://orcid.org/0000-0003-1628-4661)
:::

::: {.honor-chips}
[National Academy of Education (2025)]{.chip}
[AERA Fellow (2023)]{.chip}
[NCME President (2021–22)]{.chip}
[AERA Div. D Significant Contribution (2024)]{.chip}
[h-index 34]{.chip}
:::

:::

:::

::: {.last-updated}
*Short CV · Last updated `r format(Sys.Date(), "%B %Y")` · [Full CV →](cv.html)*
:::

## Research Focus

Psychometrics, causal inference, and assessment policy — with sustained work
on vertical scaling and growth modeling, the historical and philosophical
foundations of measurement in the human sciences, learning-progression-based
scale construction, and the design of large-scale assessment systems that
inform teaching and learning.

## Professional Experience

::: {.timeline}

2026
:   Associate Dean of Faculty, School of Education, CU Boulder

2016–
:   Director, [Center for Assessment, Design, Research and Evaluation (CADRE)](https://www.colorado.edu/cadre/), CU Boulder

2013–
:   Professor, Research and Evaluation Methodology, CU Boulder

2009–2013
:   Associate Professor, Research and Evaluation Methodology, CU Boulder

2003–2009
:   Assistant Professor, Research and Evaluation Methodology, CU Boulder

2002–2003
:   Postdoctoral Scholar, Berkeley Evaluation and Assessment Research Center, UC Berkeley

:::

## Education

::: {.timeline}

2002
:   **Ph.D.**, University of California, Berkeley — Education: Quantitative Methods and Evaluation. Advisor: Mark Wilson.

1993
:   **B.A.**, Carleton College — Economics (Departmental Distinction).

:::

## Selected Honors & Awards

::: {.timeline}

2025
:   Elected to the National Academy of Education

2024
:   Award for Significant Contribution to Educational Measurement and Research Methodology, AERA Division D

2023
:   AERA Fellow

2021–22
:   President, National Council on Measurement in Education

2020–22
:   Honorary Research Fellow, University of Oxford, Department of Education

2013–16
:   Editor, *Educational Measurement: Issues and Practice*

2012
:   NCME Annual Award for Contributions to Theory and Practice

2007–09
:   NAEd / Spencer Postdoctoral Fellowship

2004
:   AERA Division D Outstanding Dissertation Award

:::

## Selected Publications

```{r}
#| results: asis
render_featured()
```

## Grants and Contracts

::: {.grants-summary}
**Total: `r fmt_m(g_total)` across `r g_n` projects** — `r fmt_m(g_pi)` as PI
(`r g_pi_n` projects) · `r fmt_m(g_copi)` as Co-PI. Funders include the
National Science Foundation, Institute of Education Sciences, Spencer Foundation,
Carnegie Corporation, Keck Foundation, Curriculum Associates, and Denver Public
Schools.
:::

## Selected Service & Leadership

::: {.service-compact}
- **Editor**, *Educational Measurement: Issues and Practice*, 2013–2015
- **Executive Committee** (elected), National Council on Measurement in Education, 2020–2023
- **At-Large Board Member** (elected), NCME, 2016–2019
- **Chair**, Technical Advisory Committee, Multistate Alternate Assessment Consortium (since 2016)
- Technical Advisory Panels: Smarter Balanced Assessment Consortium (since 2011); PARCC / New Meridian (since 2011); NAEP / ETS Design Advisory Committee (since 2013); state TACs for AZ, KS, NY, UT
- **Editorial Boards**: *Journal of Educational Measurement* (since 2016); *Educational Assessment* (since 2010)
- **NCME Presidential Address**, 2022 — *Turning the page to the next chapter of educational measurement*
- **PhD students graduated**: 15 (2008–2025) at CU Boulder
:::
