---
title: "Derek C. Briggs — Curriculum Vitae"
author: "Derek C. Briggs"
date: last-modified
date-format: "MMMM YYYY"
format:
  html:
    theme: none
    css: cv.css
    toc: true
    toc-location: right
    toc-depth: 2
    toc-title: "Contents"
    anchor-sections: true
    link-external-newwindow: true
    include-in-header:
      text: |
        <meta name="description" content="Curriculum vitae of Derek C. Briggs, Professor of Research and Evaluation Methodology, University of Colorado Boulder.">
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(yaml)
library(readr)
library(knitr)

pubs_raw <- yaml::read_yaml("data/publications.yaml")$publications
grants_raw <- readr::read_csv("data/grants.csv", show_col_types = FALSE)
students_raw <- readr::read_csv("data/students.csv", show_col_types = FALSE)
presentations_raw <- yaml::read_yaml("data/presentations.yaml")$presentations

# -----------------------------------------------------------------
# Counts for section headings + the h-index line under Publications.
# Update h_index and metrics_date once/year (or whenever Scholar is checked).
# -----------------------------------------------------------------
h_index      <- 34L              # Google Scholar, verify at
                                 # https://scholar.google.com/citations?user=FsarJmwAAAAJ
metrics_date <- "April 2026"

n_by_category <- function(cat) sum(vapply(pubs_raw,
  function(p) identical(p$category, cat), logical(1)))

n_books            <- n_by_category("book")
n_journal_articles <- n_by_category("journal-article")
n_book_chapters    <- n_by_category("book-chapter")
n_reports          <- n_by_category("report")
n_commentaries     <- n_by_category("commentary")
n_blogs            <- n_by_category("blog")
n_in_progress      <- n_by_category("in-progress")

n_invited_talks    <- sum(vapply(presentations_raw,
  function(p) identical(p$category, "invited"), logical(1)))
n_conf_total       <- sum(vapply(presentations_raw,
  function(p) identical(p$category, "conference"), logical(1)))
n_conf_selected    <- sum(vapply(presentations_raw,
  function(p) identical(p$category, "conference") && as.numeric(p$year) >= 2015,
  logical(1)))

# Escape HTML special chars in plain strings
esc <- function(x) {
  if (is.null(x)) return("")
  x <- as.character(x)
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;", x, fixed = TRUE)
  x <- gsub(">", "&gt;", x, fixed = TRUE)
  x
}

# Sort key: integer years normal; "in press" / "in preparation" float to top.
year_sort_key <- function(y) {
  if (is.numeric(y)) return(as.numeric(y))
  if (identical(y, "in press")) return(9999)
  if (identical(y, "in preparation")) return(9998)
  suppressWarnings(as.numeric(y))
}

# Build "volume(issue), pages" tail used for journal articles & commentaries.
vol_tail <- function(p) {
  out <- ""
  if (!is.null(p$volume)) {
    out <- esc(p$volume)
    if (!is.null(p$issue)) out <- paste0(out, "(", esc(p$issue), ")")
  }
  if (!is.null(p$pages)) {
    out <- if (nchar(out)) paste0(out, ", ", esc(p$pages)) else esc(p$pages)
  }
  if (nchar(out)) paste0(", ", out) else ""
}

# Build "Location: Publisher" string.
loc_pub <- function(p) {
  if (is.null(p$publisher) && is.null(p$location)) return("")
  if (!is.null(p$location) && !is.null(p$publisher)) {
    sprintf("%s: %s", esc(p$location), esc(p$publisher))
  } else if (!is.null(p$publisher)) {
    esc(p$publisher)
  } else {
    esc(p$location)
  }
}

title_html <- function(p) esc(p$title)

# "." after title unless title already ends with ? or ! (avoids "?." double punct).
title_end <- function(p) {
  last <- substr(p$title, nchar(p$title), nchar(p$title))
  if (last %in% c("?", "!", ".")) "" else "."
}

# Linking: DOI preferred, else URL. Both get .pub-doi class.
link_html <- function(p) {
  if (!is.null(p$doi)) {
    sprintf(' <a href="https://doi.org/%s" class="pub-doi">https://doi.org/%s</a>',
            esc(p$doi), esc(p$doi))
  } else if (!is.null(p$url)) {
    sprintf(' <a href="%s" class="pub-doi">%s</a>', esc(p$url), esc(p$url))
  } else ""
}

# Per-category body.
format_pub <- function(p) {
  year_str <- as.character(p$year)
  t <- title_html(p)
  te <- title_end(p)
  a <- esc(p$authors)

  body <- switch(p$category,
    "book" = {
      lp <- loc_pub(p)
      sprintf("%s (%s). <em>%s</em>%s%s",
              a, year_str, t, te,
              if (nchar(lp)) paste0(" ", lp, ".") else "")
    },

    "journal-article" = {
      venue <- if (!is.null(p$venue)) sprintf(" <em>%s</em>%s.", esc(p$venue), vol_tail(p)) else ""
      sprintf("%s (%s). %s%s%s", a, year_str, t, te, venue)
    },

    "commentary" = {
      if (!is.null(p$editors)) {
        book <- if (!is.null(p$venue)) sprintf(" In %s, <em>%s</em>.", esc(p$editors), esc(p$venue)) else ""
        lp <- loc_pub(p); lp_str <- if (nchar(lp)) paste0(" ", lp, ".") else ""
        sprintf("%s (%s). %s%s%s%s", a, year_str, t, te, book, lp_str)
      } else {
        venue <- if (!is.null(p$venue)) sprintf(" <em>%s</em>%s.", esc(p$venue), vol_tail(p)) else ""
        sprintf("%s (%s). %s%s%s", a, year_str, t, te, venue)
      }
    },

    "book-chapter" = {
      ed <- if (!is.null(p$editors)) paste0(" In ", esc(p$editors), ",") else " In"
      book <- if (!is.null(p$venue)) sprintf(" <em>%s</em>", esc(p$venue)) else ""
      pp <- if (!is.null(p$pages)) sprintf(" (pp. %s)", esc(p$pages)) else ""
      lp <- loc_pub(p); lp_str <- if (nchar(lp)) paste0(" ", lp, ".") else ""
      sprintf("%s (%s). %s%s%s%s%s.%s",
              a, year_str, t, te, ed, book, pp, lp_str)
    },

    "report" = {
      venue <- if (!is.null(p$venue)) esc(p$venue) else ""
      loc <- if (!is.null(p$location)) esc(p$location) else ""
      tail <- if (nchar(loc) && nchar(venue)) sprintf(" %s: %s.", loc, venue)
              else if (nchar(venue)) sprintf(" %s.", venue)
              else if (nchar(loc))   sprintf(" %s.", loc)
              else ""
      sprintf("%s (%s). <em>%s</em>%s%s", a, year_str, t, te, tail)
    },

    "blog" = {
      venue <- if (!is.null(p$venue)) sprintf(" <em>%s</em>.", esc(p$venue)) else ""
      sprintf("%s (%s). %s%s%s", a, year_str, t, te, venue)
    },

    "in-progress" = {
      status <- if (!is.null(p$status)) sprintf(" (%s)", esc(p$status)) else ""
      venue <- if (!is.null(p$venue)) sprintf(" %s.", esc(p$venue)) else ""
      sprintf("%s%s. %s%s%s", a, status, t, te, venue)
    },

    sprintf("%s (%s). %s%s", a, year_str, t, te)
  )

  if (!is.null(p$note)) body <- paste0(body, " ", esc(p$note), ".")
  body <- paste0(body, link_html(p))

  sprintf('<div class="pub-entry">%s</div>', body)
}

render_category <- function(category_name) {
  entries <- Filter(function(p) identical(p$category, category_name), pubs_raw)
  if (!length(entries)) return(invisible())
  keys <- vapply(entries, function(p) year_sort_key(p$year), numeric(1))
  entries <- entries[order(keys, decreasing = TRUE)]
  base::cat(paste(vapply(entries, format_pub, character(1)), collapse = "\n\n"))
  base::cat("\n")
}

# Render books and book chapters as a single reverse-chronological list.
# Used in place of two separate sections because "Books" alone is a 1-entry
# category and looks lonely on its own; merging with chapters preserves the
# book's visibility within a unified timeline of long-form scholarly output.
render_books_and_chapters <- function() {
  entries <- Filter(function(p) p$category %in% c("book", "book-chapter"), pubs_raw)
  if (!length(entries)) return(invisible())
  keys <- vapply(entries, function(p) year_sort_key(p$year), numeric(1))
  entries <- entries[order(keys, decreasing = TRUE)]
  base::cat(paste(vapply(entries, format_pub, character(1)), collapse = "\n\n"))
  base::cat("\n")
}

# Render the small curated set of signature works flagged with `featured: true`.
# Reverse-chronological. Appears at the top of the Publications section (in the
# By Type view). Entries still appear in their normal category lists below.
render_featured <- function() {
  entries <- Filter(function(p) isTRUE(p$featured), pubs_raw)
  if (!length(entries)) return(invisible())
  keys <- vapply(entries, function(p) year_sort_key(p$year), numeric(1))
  entries <- entries[order(keys, decreasing = TRUE)]
  base::cat(paste(vapply(entries, format_pub, character(1)), collapse = "\n\n"))
  base::cat("\n")
}

# Render publications grouped by tag (topic). A publication with N tags
# appears N times — intentional duplication, since many papers sit at the
# intersection of multiple topics. Tags rendered alphabetically; within a
# tag, entries are reverse-chronological.
title_case_tag <- function(s) {
  # Preserve lowercase words that shouldn't be capitalized (e.g. "of", "and")
  small <- c("of", "and", "the", "in", "on", "for", "to", "vs")
  parts <- strsplit(s, " ")[[1]]
  out <- vapply(seq_along(parts), function(i) {
    w <- parts[i]
    if (i > 1 && tolower(w) %in% small) tolower(w)
    else paste0(toupper(substr(w, 1, 1)), substr(w, 2, nchar(w)))
  }, character(1))
  paste(out, collapse = " ")
}

render_by_topic <- function() {
  # Mirror the by-type view by excluding categories hidden there (currently
  # in-progress). Keeps the two views consistent — a publication hidden in one
  # should be hidden in the other.
  pubs_visible <- Filter(function(p) !identical(p$category, "in-progress"), pubs_raw)
  all_tags <- sort(unique(unlist(lapply(pubs_visible, function(p) p$tags))))
  all_tags <- all_tags[nzchar(all_tags)]
  for (tag in all_tags) {
    entries <- Filter(function(p) !is.null(p$tags) && tag %in% p$tags, pubs_visible)
    if (!length(entries)) next
    keys <- vapply(entries, function(p) year_sort_key(p$year), numeric(1))
    entries <- entries[order(keys, decreasing = TRUE)]
    base::cat(sprintf("\n### %s (%d)\n\n",
                      title_case_tag(tag), length(entries)))
    base::cat(paste(vapply(entries, format_pub, character(1)), collapse = "\n\n"))
    base::cat("\n")
  }
}

# --- Grants table -----------------------------------------------------------

render_grants <- function() {
  g <- grants_raw
  g <- g[order(g$end, decreasing = TRUE), ]

  fmt_range <- function(s, e) {
    sy <- substr(as.character(s), 1, 4)
    ey <- substr(as.character(e), 1, 4)
    if (sy == ey) sy else sprintf("%s–%s", sy, ey)
  }
  dates <- mapply(fmt_range, g$start, g$end)
  amount <- sprintf("$%s", formatC(g$amount, format = "d", big.mark = ","))

  # Bold "PI" entries to visually distinguish lead from collaborative roles;
  # Co-PI stays in normal weight. Manually HTML-escape the other text-bearing
  # columns since we set escape = FALSE below to allow the <strong> tag through.
  esc_html <- function(x) {
    x <- gsub("&", "&amp;", x, fixed = TRUE)
    x <- gsub("<", "&lt;", x, fixed = TRUE)
    x <- gsub(">", "&gt;", x, fixed = TRUE)
    x
  }
  role_html <- ifelse(g$role == "PI",
                      '<strong>PI</strong>',
                      esc_html(g$role))

  display <- data.frame(
    Funder = esc_html(g$funder),
    Title = esc_html(g$title),
    Role = role_html,
    Amount = amount,
    Dates = dates,
    stringsAsFactors = FALSE
  )
  knitr::kable(display,
               align = c("l", "l", "l", "r", "l"),
               format = "html",
               escape = FALSE,
               table.attr = 'class="grants-table"')
}

# --- PhD students table -----------------------------------------------------

render_students <- function() {
  s <- students_raw
  s <- s[order(s$year, decreasing = TRUE), ]
  s$position <- ifelse(is.na(s$position) | s$position == "", "—", s$position)
  display <- data.frame(
    Name = s$name,
    Year = as.character(s$year),
    `Current position` = s$position,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
  knitr::kable(display,
               align = c("l", "l", "l"),
               format = "html",
               escape = TRUE,
               table.attr = 'class="students-table"')
}

# --- Presentations ----------------------------------------------------------

format_presentation <- function(p) {
  a <- esc(p$authors)
  y <- as.character(p$year)
  t <- esc(p$title)
  te <- title_end(p)
  d <- if (!is.null(p$details)) paste0(" ", esc(p$details)) else ""
  u <- if (!is.null(p$url))
         sprintf(' <a href="%s" class="pub-doi">%s</a>', esc(p$url), esc(p$url))
       else ""
  sprintf('<div class="pub-entry">%s (%s). %s%s%s%s</div>',
          a, y, t, te, d, u)
}

render_presentations <- function(category_name, year_min = NULL) {
  entries <- Filter(function(p) identical(p$category, category_name), presentations_raw)
  if (!is.null(year_min)) {
    entries <- Filter(function(p) as.numeric(p$year) >= year_min, entries)
  }
  if (!length(entries)) return(invisible())
  keys <- vapply(entries, function(p) as.numeric(p$year), numeric(1))
  entries <- entries[order(keys, decreasing = TRUE)]
  base::cat(paste(vapply(entries, format_presentation, character(1)), collapse = "\n\n"))
  base::cat("\n")
}
```

::: {.dashboard}

::: {.dashboard-header}

[Download PDF](cv.pdf){.pdf-download-btn download="Derek C. Briggs CV.pdf"}

# Derek C. Briggs {.name}

::: {.titles}
Professor · Research and Evaluation Methodology · University of Colorado Boulder

Associate Dean of Faculty, School of Education

Director, Center for Assessment, Design, Research and Evaluation (CADRE)
:::

::: {.contact}
[derek.briggs@colorado.edu](mailto:derek.briggs@colorado.edu) ·
[Website](https://derekcbriggs.github.io/) ·
[Google Scholar](https://scholar.google.com/citations?user=FsarJmwAAAAJ&hl=en) ·
[ORCID: 0000-0003-1628-4661](https://orcid.org/0000-0003-1628-4661)
:::

::: {.honor-chips}
[National Academy of Education (2025)]{.chip}
[AERA Fellow (2023)]{.chip}
[NCME President (2021–22)]{.chip}
:::

:::

:::

::: {.last-updated}
*Last updated `r format(Sys.Date(), "%B %Y")`*
:::

## Research Expertise

My research examines how measurement works—and how it can mislead—in educational
contexts. I work at the intersection of psychometrics, causal inference, and
assessment policy, with sustained attention to vertical scaling and growth
modeling in K–12 assessment, the theoretical and historical foundations of
measurement in the human sciences, and the design of large-scale assessment
systems that can genuinely inform teaching and learning. Recent work develops
learning progressions as an alternative framework for scale construction,
content-referenced approaches to growth reporting, and the philosophical
commitments that distinguish measurement in the social sciences from its
counterparts in the physical sciences.

*Keywords:* assessment policy; causal inference in education;
educational measurement; growth modeling; history and philosophy of measurement;
item response theory; learning progressions; psychometrics; Rasch model;
vertical scaling.

## Professional Experience

::: {.timeline}

2026
:   Associate Dean of Faculty, School of Education, University of Colorado Boulder

2025
:   Visiting Scholar, Curriculum Associates

2016--present
:   Director, [Center for Assessment, Design, Research and Evaluation (CADRE)](https://www.colorado.edu/cadre/), University of Colorado Boulder

2013--present
:   Professor, Research and Evaluation Methodology, School of Education, University of Colorado Boulder

2009--2013
:   Associate Professor, Research and Evaluation Methodology, School of Education, University of Colorado Boulder

2003--2009
:   Assistant Professor, Research and Evaluation Methodology, School of Education, University of Colorado Boulder

2002--2003
:   Postdoctoral Scholar, Berkeley Evaluation and Assessment Research Center, University of California, Berkeley

1993--1996
:   Assistant Analyst, Macroeconomic Analysis Division, Congressional Budget Office, Washington, D.C.

:::

## Education

::: {.timeline}

2002
:   **Ph.D.**, University of California, Berkeley. Education: Quantitative Methods and Evaluation. Dissertation: *SAT Coaching, Bias and Causal Inference* (winner, AERA Division D Outstanding Dissertation Award, 2004). Advisor: Mark Wilson.

1993
:   **B.A.**, Carleton College. Economics (Departmental Distinction).

:::

## Honors & Awards

::: {.timeline}

2025
:   Elected to the National Academy of Education

2024
:   Award for Significant Contribution to Educational Measurement and Research Methodology, American Educational Research Association Division D, for the book *Historical and Conceptual Foundations of Measurement in the Human Sciences* (Briggs, 2021)

2023
:   American Educational Research Association Fellow

2021--2022
:   Elected President, National Council for Measurement in Education (presidential term)

2020--2022
:   Honorary Research Fellow, University of Oxford, Department of Education

2016--2019
:   Elected at-large member of Board of Directors, National Council for Measurement in Education

2013--2016
:   Editor, *Educational Measurement: Issues and Practice*

2013
:   Outstanding Reviewer Award, *Journal of Educational and Behavioral Statistics*

2012
:   University of Colorado Provost's Award for Faculty Achievement

2012
:   Annual Award for Contributions to Theory and Practice, National Council on Measurement in Education

2012
:   Outstanding Reviewer Award, *Educational Researcher*

2007--2009
:   National Academy of Education / Spencer Postdoctoral Fellowship

2004
:   AERA Division D Outstanding Dissertation Award

2003
:   UC Berkeley Graduate School of Education Commencement Address

2002
:   Educational Testing Service Summer Associate

2000
:   RAND Summer Associate

1993
:   Departmental Distinction in Economics, Carleton College

:::

## Grants and Contracts

```{r}
#| results: asis
render_grants()
```

```{r}
#| results: asis
total <- sum(grants_raw$amount)
pi_total <- sum(grants_raw$amount[grants_raw$role == "PI"])
copi_total <- total - pi_total
base::cat(sprintf(
  "\n**Totals:** $%s across %d projects (PI: $%s · Co-PI: $%s)\n",
  formatC(total, big.mark = ",", format = "d"),
  nrow(grants_raw),
  formatC(pi_total, big.mark = ",", format = "d"),
  formatC(copi_total, big.mark = ",", format = "d")
))
```

## Teaching

### Courses

- EDUC 8230: Quantitative Methods in Educational Research I
- EDUC 8710: Measurement in Survey Research
- EDUC 8720: [Psychometric Modeling: Item Response Theory](https://DerekCBriggs.github.io/Measurement-and-Psychometrics/)
- EDUC 8804: Using AI in Academic Research
- EDUC 7396: Latent Variable and Structural Equation Modeling

### PhD Students Graduated

```{r}
#| results: asis
render_students()
```

## Publications

::: {.section-meta}
**h-index:** `r h_index` &nbsp;·&nbsp; Google Scholar, `r metrics_date`
:::

<div class="pub-view-toggle" role="tablist" aria-label="Publication grouping">
<button type="button" class="active" data-view="by-type" aria-pressed="true">By Type</button>
<button type="button" data-view="by-topic" aria-pressed="false">By Topic</button>
</div>

::: {.pub-view data-view="by-type"}

### Selected Publications

```{r}
#| results: asis
render_featured()
```

### Refereed Journal Articles (`r n_journal_articles`)

```{r}
#| results: asis
render_category("journal-article")
```

### Books and Book Chapters (`r n_books + n_book_chapters`)

```{r}
#| results: asis
render_books_and_chapters()
```

### Reports & Working Papers (`r n_reports`)

```{r}
#| results: asis
render_category("report")
```

### Commentaries & Reviews (`r n_commentaries`)

```{r}
#| results: asis
render_category("commentary")
```

### Blogs, Interviews, Podcasts (`r n_blogs`)

```{r}
#| results: asis
render_category("blog")
```

:::

::: {.pub-view data-view="by-topic" hidden="hidden"}

```{r}
#| results: asis
render_by_topic()
```

:::

## Presentations

### Invited Talks & Keynotes (`r n_invited_talks`)

```{r}
#| results: asis
render_presentations("invited")
```

### Conference Presentations since 2015 (`r n_conf_selected`)

```{r}
#| results: asis
render_presentations("conference", year_min = 2015)
```

[See full list of `r n_conf_total` conference presentations →](presentations-full.html){.pdf-download}

## Professional Service

### National & International Leadership

- Executive Committee (elected), National Council for Measurement in Education, 2020--2023
- At-Large Board Member (elected), National Council for Measurement in Education, 2016--2019
- Research Methods Section Chair, Society for Research on Educational Effectiveness (SREE), Spring 2014 Conference
- NCME Dissertation Award Committee Chair, 2010--2011
- Conference Chair, International Objective Measurement Workshop, 2010
- AERA Division L, Section 5 Program Chair, 2009--2010

### Editorial Roles

#### Editor

- Editor, *Educational Measurement: Issues and Practice*, 2013--2015

#### Editorial Boards (Current)

- *Educational Assessment* (since 2010)
- *Journal of Educational Measurement* (since 2016)

#### Editorial Boards (Past)

- *AERA Open* (2014--2021)
- *International Journal of Testing* (2016--2019)
- *Educational Measurement: Issues and Practice* (2007--2009)
- *Educational Researcher* (2006--2009)

### Advisory Panels & Technical Committees

#### Current

- ETS-Carnegie Job for the Future Panel (since 2024)
- NAEP Validity Studies Panel (since 2023)
- Curriculum Associates, Technical Advisory Committee (since 2018)
- Multistate Alternate Assessment Consortium, Chair of Technical Advisory Committee (since 2016)
- State of Kansas Assessment Technical Advisory Committee (since 2019)
- State of Arizona Assessment Technical Advisory Committee (since 2016)
- National Assessment of Educational Progress, ETS Design Advisory Committee (since 2013)
- Partnership for the Assessment of College and Career Readiness Consortium (now New Meridian), Technical Advisory Panel (since 2011)
- Smarter Balanced Assessment Consortium (SBAC), Technical Advisory Panel (since 2011)
- State of New York Assessment Technical Advisory Committee (since 2012)
- State of Utah Assessment Technical Advisory Committee (since 2007)

#### Past

- Educational Testing Service Visiting Panel on Research (2019--2025)
- NAEP Reading Framework Revision, Technical Advisory Panel (2019--2021)
- NAEP Mathematics Framework Revision, Technical Advisory Panel (2018--2019)
- Institute of Education Sciences Review Panel, Education Systems and Broad Reform (2016--2019)
- State of Nevada Assessment Technical Advisory Committee (2016--2019)
- State of Indiana Assessment Technical Advisory Committee (2016)
- State of Tennessee Assessment Technical Advisory Committee (2015--2018)
- State of Michigan Assessment Technical Advisory Committee (2013--2015)
- English Language Proficiency Assessment Consortium (ELPA21), Technical Advisory Committee (2013--2016)
- National Centers and State Collaborative (NCSC), Technical Advisory Committee (2011--2015)
- Georgia Educator Effectiveness Technical Advisory Committee (2011--2019)
- State of Montana Assessment Technical Advisory Committee (2006--2017)
- Gates Foundation, Measures of Effective Teaching Technical Advisory Panel (2010)
- National Council on Measurement in Education, Brenda Loyd Dissertation Award Committee (2008--2011)
- Expert Panel, Evaluating the Validity of English Language Proficiency Assessments (2009--2011)
- Expert Panel, GSEG Consortia: Validity Evaluation, National Alternate Assessment Center (2008--2009)
- Expert Panel, Growth Model Task Force, National Center for Learning Disabilities (2008)

### Institutional Service — University of Colorado Boulder

- Associate Dean of Faculty, School of Education, 2026--
- Faculty Salary Equity Fellow, Office of the Provost, 2025--2026
- Vice-Chancellor's Advisory Committee (Promotion and Tenure Review), 2024
- School of Education, Faculty Merit Review Committee, 2023--2025
- School of Education, Research and Evaluation Methodology Program Chair, 2008--2019
- Chair, Research and Evaluation Methodology Search Committees, 2008--2016 (multiple cycles)
- Member, Dean Search Committee, 2014--2016
- Coordinator of Bi-Weekly Research & Evaluation Methodology Seminar, 2003--2014
- Graduate Student Recruitment Taskforce, 2006--2008
- Science Education Search Committee, 2005--2006
- Doctoral Curriculum Taskforce, 2003--2005

### Manuscript Reviewing

::: {.manuscript-reviews}
- *AERA Open*
- *American Educational Research Journal*
- *Applied Psychological Measurement*
- *Assessment in Education*
- *Behavioral Research Methods*
- *British Journal of Mathematical and Statistical Psychology*
- *Cognition and Instruction*
- *Educational Evaluation and Policy Analysis*
- *Educational Measurement: Issues and Practice*
- *Educational Policy*
- *Educational Researcher*
- *Evaluation Review*
- *International Journal of Testing*
- *Journal of Educational and Behavioral Statistics*
- *Journal of Educational Measurement*
- *Journal of Experimental Child Psychology*
- *Journal of Teacher Education*
- *Multivariate Behavioral Research*
- *Physical Review*
- *Psychological Methods*
- *Psychometrika*
- *Review of Educational Research*
- *Theory & Psychology*
- Routledge Publications
- SAGE Publications
:::

### Professional Affiliations

National Academy of Education; American Educational Research Association; National Council on Measurement in Education; The Psychometric Society; Society for the Study of Measurement; National Education Policy Center.

<script>
// Publication view toggle: swap between "By Type" and "By Topic" groupings.
(function () {
  const toggle = document.querySelector('.pub-view-toggle');
  if (!toggle) return;
  const views = document.querySelectorAll('.pub-view');
  toggle.addEventListener('click', function (e) {
    const btn = e.target.closest('button[data-view]');
    if (!btn) return;
    const target = btn.dataset.view;
    toggle.querySelectorAll('button').forEach(function (b) {
      const isActive = b.dataset.view === target;
      b.classList.toggle('active', isActive);
      b.setAttribute('aria-pressed', isActive ? 'true' : 'false');
    });
    views.forEach(function (v) {
      if (v.dataset.view === target) v.removeAttribute('hidden');
      else v.setAttribute('hidden', 'hidden');
    });
  });
})();

// Prepend a "Derek C. Briggs" entry to the TOC that returns to the dashboard.
// Quarto's TOC starts at h2 by design, so the dashboard h1 is excluded unless
// we inject it. We look up the dashboard section's auto-generated ID at load
// time so the anchor stays correct even if the heading text changes.
(function () {
  const tocList = document.querySelector('nav#TOC > ul');
  const dashSection = document.querySelector('section.dashboard-header');
  if (!tocList || !dashSection || !dashSection.id) return;
  const li = document.createElement('li');
  const a = document.createElement('a');
  a.href = '#' + dashSection.id;
  a.textContent = 'Derek C. Briggs';
  li.appendChild(a);
  tocList.insertBefore(li, tocList.firstChild);
})();
</script>
